\documentclass[10pt,twocolumn]{article}

\usepackage[margin=0.68in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage{enumitem}
\usepackage[table]{xcolor}
\usepackage{hyperref}
\usepackage{titlesec}
\usepackage{graphicx}
\usepackage[section]{placeins}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric,arrows.meta,positioning,fit,backgrounds,calc}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!45!black,
  urlcolor=blue!45!black,
  citecolor=blue!45!black,
  pdftitle={Long-Term Memory for Coding Agents},
  pdfauthor={Cortex Project}
}

\setlength{\columnsep}{0.24in}
\setlength{\parindent}{0.9em}
\setlength{\parskip}{0pt}
\setlength{\emergencystretch}{1.25em}
\Urlmuskip=0mu plus 1mu
\linespread{0.98}
\setlist{itemsep=2pt,topsep=2pt,leftmargin=1.2em}
\titlespacing*{\section}{0pt}{0.9ex plus .2ex minus .2ex}{0.5ex}
\titlespacing*{\subsection}{0pt}{0.7ex plus .2ex minus .2ex}{0.4ex}
\titlespacing*{\subsubsection}{0pt}{0.6ex plus .2ex minus .2ex}{0.3ex}

% Scorecard colors: H = green (good), M = amber (mid), L = red (bad)
% For Lock-in column: L (low lock-in) = good, H (high lock-in) = bad
% We handle this by using separate color macros per column in the table
\definecolor{scoreHi}{HTML}{CFEAD6}   % green  -- high score, good
\definecolor{scoreMed}{HTML}{F5E8C8}  % amber  -- medium
\definecolor{scoreLow}{HTML}{F6CDCD}  % red    -- low score, bad
\definecolor{scoreLkin}{HTML}{F6CDCD} % red    -- HIGH lock-in (bad)
\definecolor{scoreLklo}{HTML}{CFEAD6} % green  -- LOW lock-in (good)
\definecolor{scoreDark}{HTML}{134E4A}

\title{\vspace{-1.2em}Long-Term Memory for Coding Agents\\
\large Architecture, Governance, Multi-Agent Coordination, and Comparative Evaluation}
\author{Cortex Project}
\date{March 8, 2026 \quad\textbar\quad v1.15.5}

\begin{document}
\maketitle

\begin{abstract}
Persistent memory is a first-order requirement for multi-session coding workflows. This report documents why Cortex chose a markdown-native, git-backed architecture with hook-driven lifecycle automation, trust-scored retrieval, and configurable governance. Cortex v1.15.5 ships a policy engine (retention, decay, access control), an interactive shell for direct memory management, index-level filtering, user-extensible synonym configuration, cross-project age-decay ranking, per-project skill injection controls, and a skills/hooks management CLI. The report compares Cortex to Supermemory, claude-mem, and GitHub Copilot Memory, provides a cost model with worked examples, and proposes a reproducible benchmark protocol. The core argument is operational: in real engineering teams, memory quality has to be optimized alongside auditability, governance, portability, and token predictability.
\end{abstract}

\section{Executive Summary}
A memory system for coding agents must solve four coupled problems:
\begin{enumerate}
  \item preserve context over time,
  \item retrieve relevant context under token constraints,
  \item suppress stale or low-confidence memory,
  \item remain operable at team scale.
\end{enumerate}

Cortex optimizes for ownership and controllability. Alternatives can outperform on managed convenience, setup simplicity, or deep platform integration. The right choice depends on which constraints are binding.
This report separates \textit{architecture judgments} from \textit{measured performance}; they are not interchangeable. Sections without empirical data say so directly.

\section{Decision Criteria}
This report evaluates systems on eight dimensions:
\begin{enumerate}
  \item \textbf{Representation}: where memory lives and who owns it.
  \item \textbf{Capture path}: how memory is created (hooks/API/workers).
  \item \textbf{Retrieval path}: ranking, trust scoring, and injection under budget.
  \item \textbf{Governance}: forgetting, validation, review, and policy controls.
  \item \textbf{Economics}: token overhead + human maintenance burden.
  \item \textbf{Portability}: migration risk and lock-in profile.
  \item \textbf{Multi-agent readiness}: concurrent sessions, shared state, coordination.
  \item \textbf{Operability}: CLI, shell, direct management without the model in the loop.
\end{enumerate}

\section{Cortex Architecture}

The system has three primary subsystems: \textbf{Storage \& Retrieval}, \textbf{Lifecycle Automation}, and \textbf{Governance}. Figure~\ref{fig:arch} shows how they connect.

\begin{figure*}[!t]
\centering
\begin{tikzpicture}[
  font=\small,
  node distance=0.55cm and 0.9cm,
  box/.style={rectangle, rounded corners=3pt, draw, fill=blue!8, minimum height=1.9em, align=center, inner sep=4pt},
  hook/.style={rectangle, rounded corners=3pt, draw, fill=orange!15, minimum height=1.9em, align=center, inner sep=4pt},
  store/.style={rectangle, rounded corners=3pt, draw, fill=green!12, minimum height=1.9em, align=center, inner sep=4pt},
  gov/.style={rectangle, rounded corners=3pt, draw, fill=purple!10, minimum height=1.9em, align=center, inner sep=4pt},
  arr/.style={-{Stealth[length=5pt]}, thick},
  dasharr/.style={-{Stealth[length=5pt]}, dashed, gray},
]

% === INPUT ===
\node[box] (prompt) {User Prompt};

% === HOOKS ===
\node[hook, right=1.1cm of prompt] (ss) {SessionStart\\[-2pt]{\scriptsize git pull}};
\node[hook, below=0.55cm of ss]    (ups) {UserPromptSubmit\\[-2pt]{\scriptsize keyword extract, inject}};
\node[hook, below=0.55cm of ups]   (stop) {Stop\\[-2pt]{\scriptsize git commit + push}};

% === RETRIEVAL ===
\node[box, right=1.1cm of ups] (fts) {FTS5 Index\\[-2pt]{\scriptsize SQLite, cached snapshot}};
\node[box, above=0.55cm of fts] (syn)  {Synonym Expansion\\[-2pt]{\scriptsize utils.ts}};
\node[box, below=0.55cm of fts] (trust) {Trust Filter\\[-2pt]{\scriptsize age decay, citation score}};

% === STORAGE ===
\node[store, right=1.1cm of fts] (md) {Markdown Files\\[-2pt]{\scriptsize \textasciitilde/.cortex/<project>/}};
\node[store, above=0.55cm of md]  (git) {Git History\\[-2pt]{\scriptsize audit + portability}};
\node[store, below=0.55cm of md]  (mcp) {MCP Tools (40)\\[-2pt]{\scriptsize add\_finding, backlog, skills, hooks, etc.}};

% === GOVERNANCE ===
\node[gov, right=1.1cm of md] (pol) {Policy Engine\\[-2pt]{\scriptsize .governance/*.json}};
\node[gov, above=0.55cm of pol] (rbac) {Access Control\\[-2pt]{\scriptsize 4 roles, 6 actions}};
\node[gov, below=0.55cm of pol] (aud) {Audit Log\\[-2pt]{\scriptsize memory-usage.log}};

% === ARROWS: prompt -> hooks ===
\draw[arr] (prompt) -- (ss);
\draw[arr] (prompt) -- (ups);
\draw[arr] (ups)    -- (stop);

% === ARROWS: hooks <-> retrieval ===
\draw[arr] (ups) -- (syn);
\draw[arr] (syn) -- (fts);
\draw[arr] (fts) -- (trust);
\draw[arr] (trust) -- (ups);

% === ARROWS: retrieval <-> storage ===
\draw[arr] (fts) -- (md);
\draw[arr] (md)  -- (git);
\draw[arr] (mcp) -- (md);
\draw[arr] (stop) -- (md);

% === ARROWS: storage <-> governance ===
\draw[arr] (md)  -- (pol);
\draw[arr] (pol) -- (rbac);
\draw[arr] (pol) -- (aud);

% === STARTUP: ss -> fts cache warm / context flow ===
\draw[dasharr] (ss) to[bend left=20] node[above, font=\scriptsize]{context + cache reuse} (fts);

\end{tikzpicture}
\caption{Cortex data flow. Retrieval remains markdown-native, but the local index is cached and loaded by hook and MCP entrypoints rather than rebuilt once at session start. Every prompt cycle runs: keyword extraction \textrightarrow\ FTS5 search \textrightarrow\ trust filter \textrightarrow\ context injection. Writes go through MCP tools or the Stop hook, committing to git. Permission checks gate writes; workflow approval is enforced on queue-based review operations rather than every write path.}
\label{fig:arch}
\end{figure*}

\subsection{Storage and Retrieval}
Cortex keeps memory in project markdown files and builds a local SQLite FTS5 index for retrieval \cite{cortex-readme,cortex-shared,cortex-cli}. Markdown files remain the single source of truth; the index is a cached snapshot that hook and MCP entrypoints load and incrementally refresh as files change. This yields strong inspectability (diffs, git history, plain-text grep) and low platform lock-in.

A configurable index policy \cite{cortex-shared,cortex-index} controls which files are indexed. Operators set \texttt{includeGlobs} (default: \texttt{**/*.md}), \texttt{excludeGlobs} (default: \texttt{node\_modules}, \texttt{.git}), and an \texttt{includeHidden} flag. The policy is hashed into the index cache key, so changing it forces a clean rebuild and prevents accidental indexing of vendor files or sensitive directories.

Before injection, a trust filter scores each learning entry by age decay and citation validity (Section~\ref{sec:decay}), then suppresses entries below \texttt{minInjectConfidence=0.35}. Archived entries inside \texttt{<details>} blocks are stripped so consolidated history does not pollute active search. In the current implementation, trust filtering may also queue stale or conflicting findings for later human review.

\subsection{Lifecycle Automation}
Three hook stages automate retrieval and persistence \cite{cortex-cli,cortex-hooks}:
\begin{enumerate}
  \item \texttt{SessionStart}: pulls the cortex repo and runs lightweight startup checks.
  \item \texttt{UserPromptSubmit}: extracts keywords, expands synonyms, loads or refreshes the cached local index, searches it, and injects relevant snippets within the token budget ($\approx$550 tokens, 6 lines, 520 chars per snippet). It also checks consolidation thresholds and fires a one-per-session notice when the project has accumulated $\geq$25 new findings.
  \item \texttt{Stop}: auto-commits and pushes all cortex changes after every response, including subagent responses.
\end{enumerate}
This cuts manual tool invocations and keeps session behavior consistent.

\subsection{MCP Context Optimization}
\label{sec:context-opt}

MCP tool responses consume context tokens. A full backlog fetch can return $\sim$10k tokens, which is 5\% of a 200k context window from a single tool call. For multi-step agent workflows with many tool calls, uncontrolled fetches compound quickly.

Cortex addresses this with progressive disclosure at two levels: the automatic hook path and the MCP tool path.

\textbf{Hook-level progressive disclosure.} When \texttt{CORTEX\_FEATURE\_PROGRESSIVE\_DISCLOSURE=1}, the \texttt{UserPromptSubmit} hook switches strategy based on result count. For 1--2 results, full snippets are injected directly (small enough that overhead is negligible). For 3+ results, the hook emits a compact memory index: one line per result with a \texttt{mem:project/filename} ID and a one-line summary truncated at 80 characters. The agent then calls \texttt{get\_memory\_detail(id)} to expand only the entries it needs. This keeps automatic injection small and predictable while still surfacing all relevant matches.

\textbf{Backlog progressive disclosure.} The \texttt{get\_backlog} tool supports three retrieval tiers:
\begin{enumerate}
  \item \textbf{Summary} (\texttt{summary:true}): counts and titles only, $\sim$200 tokens. Use for planning and status checks.
  \item \textbf{Paginated} (\texttt{limit}, \texttt{offset}): full item content with pagination, $\sim$1k tokens per page. Use when reviewing item details.
  \item \textbf{Single item} (\texttt{id:"bid:XXXXXXXX"}): one item by stable hash, $\sim$100 tokens. Use during execution.
\end{enumerate}

\textbf{Stable backlog identifiers.} Every backlog item embeds a content-addressed 8-character hex hash as an HTML comment (\texttt{<!-- bid:a3f9c2e1 -->}). These hashes survive reordering, completions, and edits to other items. Positional IDs (A1, Q3, D2) shift when items move between sections; stable \texttt{bid:} hashes do not. For multi-agent coordination, pass \texttt{bid:} hashes in task descriptions so downstream agents can fetch their specific item with \texttt{get\_backlog(id:"bid:xxxx")} instead of scanning the full backlog.

The general principle: default to the smallest fetch that answers the current question. Expand only when the agent needs more detail to proceed. This applies equally to \texttt{get\_findings}, \texttt{search\_knowledge}, and any other tool that returns variable-length content.

\subsection{Governance}
\label{sec:governance}

The governance engine lives in \texttt{.governance/} \cite{cortex-shared} and is composed of four policy files:

\textbf{Retention policy} (\texttt{retention-policy.json}): TTL (\texttt{ttlDays=120}), retention (\texttt{retentionDays=365}), auto-accept confidence threshold, minimum injection confidence, and the four-tier decay curve. Changes are audit-logged.

\textbf{Workflow policy} (\texttt{workflow-policy.json}): approval gates controlling review-queue approval of risky memories and which queue sections trigger maintainer review.

\textbf{Access control} (\texttt{access-control.json}): role-based permissions with four roles: \texttt{admin}, \texttt{maintainer}, \texttt{contributor}, \texttt{viewer}. Each maps to a set of allowed actions (\texttt{read}, \texttt{write}, \texttt{queue}, \texttt{pin}, \texttt{policy}, \texttt{delete}). The system resolves the current actor's role and rejects unauthorized operations before they execute.

\textbf{Canonical locks} (\texttt{canonical-locks.json}): pinned memories are protected by content hashes. This prevents accidental overwrite during consolidation or bulk operations.

Governance and memory-quality events are appended to \texttt{memory-usage.log} with timestamps, providing a human-readable audit trail without requiring database queries.

\subsection{Trust Decay and Confidence Scoring}
\label{sec:decay}

Not all memories deserve equal weight. Each learning entry is scored before injection:
\begin{itemize}
  \item \textbf{Age decay}: confidence degrades over four configurable windows (\texttt{d30}/\texttt{d60}/\texttt{d90}/\texttt{d120}), defaulting to 1.0, 0.9, 0.75, and 0.5 respectively.
  \item \textbf{Citation validity}: entries with valid file/line/commit citations score higher. Entries without citations receive a 0.8 multiplier.
  \item \textbf{Minimum confidence}: entries below \texttt{minInjectConfidence=0.35} are suppressed.
  \item \textbf{Quality feedback}: impressions, helpful-count, reprompt penalty, and regression penalty are tracked per entry. Entries that cause regressions lose injection priority over time.
\end{itemize}

Figure~\ref{fig:decay} visualizes the four-tier decay curve and the effect of missing citations.

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=\columnwidth,
  height=4.8cm,
  xlabel={Age (days)},
  ylabel={Confidence},
  xmin=0, xmax=135,
  ymin=0, ymax=1.08,
  xtick={0,30,60,90,120},
  ytick={0,0.35,0.5,0.75,0.9,1.0},
  yticklabel style={font=\scriptsize},
  xticklabel style={font=\scriptsize},
  xlabel style={font=\small},
  ylabel style={font=\small},
  grid=both,
  grid style={gray!20},
  legend style={font=\scriptsize, at={(0.98,0.98)}, anchor=north east},
  clip=true,
]
% Cited entries (full score)
\addplot[blue!70!black, thick, mark=*, mark size=2pt] coordinates {
  (0,  1.0)
  (30, 1.0)
  (60, 0.9)
  (90, 0.75)
  (120,0.5)
};
\addlegendentry{Cited entry}

% Uncited entries (0.8x multiplier)
\addplot[orange!80!black, thick, dashed, mark=square*, mark size=2pt] coordinates {
  (0,  0.8)
  (30, 0.8)
  (60, 0.72)
  (90, 0.60)
  (120,0.40)
};
\addlegendentry{Uncited entry ($\times 0.8$)}

% Suppression threshold
\addplot[red!70!black, dotted, very thick] coordinates {(0,0.35)(135,0.35)};
\addlegendentry{Suppression floor (0.35)}

\end{axis}
\end{tikzpicture}
\caption{Trust decay curve. Cited entries hold full confidence for 30 days, then step down at 60/90/120 days. Uncited entries start at 0.8$\times$. Entries below 0.35 are never injected. Defaults; all thresholds are configurable.}
\label{fig:decay}
\end{figure}

\subsection{Extensibility}

Beyond the three lifecycle hooks, Cortex supports user-defined hooks for seven events: \texttt{pre-save}, \texttt{post-save}, \texttt{post-search}, \texttt{pre-finding}, \texttt{post-finding}, \texttt{pre-index}, and \texttt{post-index} \cite{cortex-hooks}. Hooks receive \texttt{CORTEX\_PATH} and \texttt{CORTEX\_HOOK\_EVENT} as environment variables, plus event-specific data. Typical uses include Slack notifications on new findings, webhook calls after index rebuilds, and pre-save validation scripts.

Projects can be exported as self-contained JSON bundles via \texttt{export\_project} and re-imported with \texttt{import\_project}. The \texttt{manage\_project} tool supports archiving inactive projects without deleting data. Migration tooling (\texttt{cortex maintain migrate <project>}) scans git history and GitHub PR reviews, CI failures, and issue comments, routing extracted candidates to \texttt{FINDINGS.md} or \texttt{MEMORY\_QUEUE.md} based on confidence.

\section{Interactive Shell}
The terminal-native shell (\texttt{cortex shell}) provides direct access to memory operations without requiring an LLM in the loop \cite{cortex-cli}. Eight views are navigable by single keystroke: Projects, Backlog, Findings, Review Queue, Skills, Hooks, Machines/Profiles, and Health diagnostics.

A palette system (\texttt{:command}) supports structured operations: \texttt{:add}, \texttt{:complete}, \texttt{:move}, \texttt{:govern}, \texttt{:consolidate}, \texttt{:find add}, \texttt{:mq approve}, and more. Filtering (\texttt{/keyword}) and pagination (\texttt{:page next}) handle large datasets. State persists across invocations (current view, project, page, filter) so engineers return to where they left off.

The design goal is operational: memory governance should not require starting a coding session. An engineer should be able to run \texttt{cortex shell}, review what accumulated overnight, and clean it up before any agent touches it.

\section{Multi-Agent Coordination}
Modern coding workflows frequently involve multiple agents running concurrently. Memory systems that assume a single serial session break under this pattern.

Cortex handles multi-agent use through three mechanisms. First, \textbf{shared git-backed state}: all agents read and write the same \texttt{\textasciitilde/.cortex} directory. Concurrent reads are safe at the file system layer. Within a single process, MCP writes are serialized and guarded by file locks; across separate agent processes or machines, final synchronization still relies on git pull/push behavior.

Second, \textbf{skill-driven coordination}: Cortex skills (markdown files in \texttt{global/skills/} or project-local skill directories) define reusable agent behaviors and can be injected per project. This gives teams a way to coordinate conventions and memory workflows across agents without hard-coding them into the runtime.

Third, \textbf{hook isolation}: each agent session gets its own hook execution context. The consolidation notice fires once per session (tracked via session marker files), preventing duplicate notices within a session.

\section{API vs MCP vs Local}
These concepts are often conflated:
\begin{itemize}
  \item \textbf{MCP} is an integration protocol (tooling interface).
  \item \textbf{Local vs API} is a storage/compute decision (where memory runs).
\end{itemize}

Cortex is \textit{local memory + MCP interface + lifecycle hooks}. A local system can still expose MCP tools. The current MCP surface exposes 40 tools across search, backlog management, finding capture, entity graph, session management, memory quality feedback, hooks, skills, and data portability \cite{cortex-index}.

\section{Comparative Assessment}
\textit{Evidence window: public documentation/changelogs available as of March 6, 2026.}

\subsection{System snapshots}
\begin{itemize}
  \item \textbf{Supermemory}: API-first managed memory, metadata/tag partitioning, graph-memory semantics, and benchmark tooling \cite{sm-intro,sm-create,sm-filter,sm-graph,sm-bench}.
  \item \textbf{claude-mem}: hook-driven local architecture with SQLite/FTS5 and optional semantic layer \cite{cm-overview,cm-hooks,cm-db,cm-search}.
  \item \textbf{Copilot Memory}: repository-scoped managed memory with policy controls and expiry semantics, with active rollout changes in 2026 \cite{gh-memory-docs,gh-changelog-jan2026,gh-changelog-mar2026}.
\end{itemize}

\subsection{Known figures used in this report}
\begin{itemize}
  \item Cortex injection defaults: token budget $\approx$550, snippet lines 6, snippet chars 520 \cite{cortex-cli,cortex-index}.
  \item Governance defaults: \texttt{ttlDays=120}, \texttt{retentionDays=365} \cite{cortex-shared}.
  \item Trust filter defaults: decay curve 1.0/0.9/0.75/0.5 at 30/60/90/120 days, \texttt{minInjectConfidence=0.35}, citation-absent multiplier 0.8 \cite{cortex-shared}.
  \item Access control: 4 roles (admin, maintainer, contributor, viewer), 6 action types \cite{cortex-shared}.
  \item MCP tool surface: 40 tools across search, backlog, finding capture, entity graph, session management, skills, hooks, portability, and quality feedback categories \cite{cortex-index}.
  \item Interactive shell: 8 views, palette commands, persistent state \cite{cortex-shell}.
  \item Claude Opus 4.6 includes a 1M-token context window (beta) \cite{anthropic-opus46}.
\end{itemize}
No comparable public p95 coding-latency benchmark exists across all systems/models; cross-system speed claims must be measured directly.

\subsection{Scorecard (architecture-level, non-empirical)}

Column definitions: \textbf{Ease} = zero-to-working setup friction (H=easy). \textbf{Token ctrl} = ability to bound injected tokens (H=tight control). \textbf{Governance} = policy engine, roles, approval workflows. \textbf{Audit} = queryable history of memory changes. \textbf{Trust} = confidence scoring and decay. \textbf{Multi-agent} = concurrent session safety. \textbf{Shell/CLI} = model-free management tool. \textbf{Daily ops} = ongoing maintenance burden after setup (H=low burden). \textbf{Lock-in} = data portability risk (H=high lock-in, color-coded red; L=low lock-in, color-coded green).

\begin{table*}[!t]
\centering
\scriptsize
\setlength{\tabcolsep}{3pt}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lccccccccc}
\toprule
\textbf{System} & \textbf{Ease} & \textbf{Token ctrl} & \textbf{Governance} & \textbf{Audit} & \textbf{Trust} & \textbf{Multi-agent} & \textbf{Shell/CLI} & \textbf{Daily ops} & \textbf{Lock-in} \\
\midrule
Cortex 1.15.5   & \cellcolor{scoreMed}M & \cellcolor{scoreHi}H & \cellcolor{scoreHi}H & \cellcolor{scoreHi}H & \cellcolor{scoreHi}H & \cellcolor{scoreMed}M & \cellcolor{scoreHi}H & \cellcolor{scoreMed}M & \cellcolor{scoreLklo}L \\
Supermemory     & \cellcolor{scoreHi}H  & \cellcolor{scoreMed}M & \cellcolor{scoreMed}M & \cellcolor{scoreMed}M & \cellcolor{scoreMed}M & \cellcolor{scoreLow}L & \cellcolor{scoreLow}L & \cellcolor{scoreHi}H  & \cellcolor{scoreLkin}H \\
claude-mem      & \cellcolor{scoreMed}M & \cellcolor{scoreMed}M & \cellcolor{scoreMed}M & \cellcolor{scoreMed}M & \cellcolor{scoreMed}M & \cellcolor{scoreMed}M & \cellcolor{scoreLow}L & \cellcolor{scoreMed}M & \cellcolor{scoreMed}M \\
Copilot Memory  & \cellcolor{scoreHi}H  & \cellcolor{scoreMed}M & \cellcolor{scoreMed}M & \cellcolor{scoreMed}M & \cellcolor{scoreLow}L & \cellcolor{scoreLow}L & \cellcolor{scoreLow}L & \cellcolor{scoreHi}H  & \cellcolor{scoreLkin}H \\
\bottomrule
\end{tabular}
}
\caption{Architecture-level scorecard from documented behavior (H/M/L). H=high, M=medium, L=low. For all columns except Lock-in: green=H (better), red=L (worse). For Lock-in: green=L (less lock-in, better), red=H (more lock-in, worse). This is an operator-judgment summary, not a substitute for empirical benchmark results.}
\label{tab:scorecard}
\end{table*}

\subsection{Operational Differentiators}

The systems in this comparison differ less in whether they can retrieve a relevant memory at all and more in how they behave operationally once memory becomes a shared, long-lived artifact.

For Cortex, the main operational differentiators are:

\begin{itemize}
  \item \textbf{Git-backed auditability}: every memory write is a commit. You can \texttt{git blame} any learning, revert a bulk operation, or audit who added what and when. Managed API systems have no equivalent.
  \item \textbf{Policy and access governance}: role-based access control, approval workflows for risky queue items, TTL and retention policy, and canonical locks are first-class features.
  \item \textbf{Trust filtering}: citation decay, confidence scores, and regression penalties suppress stale or low-quality memories before injection.
  \item \textbf{CLI shell for human review}: \texttt{cortex shell} provides a model-free interface for reviewing, approving, and cleaning memories before any agent consumes them. This is operationally essential at team scale.
\end{itemize}

Any memory system can add a vector index. Governance, auditability, and operational tooling require deliberate architectural choices that are harder to retrofit. This is where Cortex is differentiated most clearly.

\section{Speed and Cost Analysis}

\subsection{Latency decomposition}
End-to-end latency decomposes into five stages:
\begin{enumerate}
  \item retrieval/index lookup,
  \item ranking/trust filtering,
  \item prompt augmentation,
  \item model generation,
  \item post-response persistence and governance.
\end{enumerate}

\subsection{Operational cost equation}
\begin{align*}
\mathrm{Total\ Cost} &= \mathrm{Token\ Cost} + \mathrm{Infra\ Cost}\\
&\quad + (\mathrm{Engineer\ Hours} \times \mathrm{Loaded\ Rate})
\end{align*}

\begin{align*}
\mathrm{Token\ OpEx}_{month} &= N \times ((T_{in}+T_{inj}) \times p_{in}\\
&\quad + T_{out} \times p_{out})
\end{align*}

In production teams, the labor term is routinely underestimated. Governance cleanup and workflow drift often dominate once basic retrieval works.

\subsection{Worked cost example}

Consider a team running 500 prompts/month against Claude Opus 4.6 (input \$15/M tokens, output \$75/M tokens as of March 2026). Assume 1,500 tokens average input, 800 tokens average output, and Cortex injecting its default 550-token budget per prompt.

\begin{align*}
T_{in,total} &= 1{,}500 + 550 = 2{,}050 \text{ tokens/prompt}\\
\mathrm{Token\ Cost} &= 500 \times (2{,}050 \times \tfrac{15}{10^6} + 800 \times \tfrac{75}{10^6})\\
&= 500 \times (0.03075 + 0.060)\\
&= 500 \times 0.09075 \approx \$45.38/\mathrm{mo}
\end{align*}

Without memory injection (no Cortex overhead):
\begin{align*}
\mathrm{Token\ Cost}_{baseline} &= 500 \times (1{,}500 \times \tfrac{15}{10^6} + 800 \times \tfrac{75}{10^6})\\
&\approx \$41.25/\mathrm{mo}
\end{align*}

\textbf{Memory overhead: \$4.13/mo} at 500 prompts. At 5,000 prompts/month, the overhead is \$41.25 --- still modest relative to the value of reduced repetition and context re-entry. Two engineer-hours of re-explanation prevented per month (at a loaded rate of \$100/hr) makes the economics straightforward.

The labor and governance terms are harder to model but often larger. A team spending 30 minutes/week on memory triage pays roughly \$200/mo in loaded time --- which argues for governance automation over manual review.

\subsection{Practical measurement panel}
\begin{table}[h]
\centering
\small
\begin{tabularx}{\columnwidth}{>{\raggedright\arraybackslash}p{0.42\columnwidth} >{\raggedright\arraybackslash}X}
\toprule
\textbf{Metric} & \textbf{Interpretation} \\
\midrule
Injected tokens / prompt & Memory overhead on context window and cost \\
Task success rate & End-user outcome quality \\
Citation validity rate & Grounding and factual confidence \\
Latency p95 & Tail performance under realistic load \\
Stale-memory incidents & Quality pressure from aging memory \\
Manual interventions / 100 tasks & Governance burden and operator effort \\
\bottomrule
\end{tabularx}
\caption{Minimum panel for meaningful speed/cost/quality analysis.}
\end{table}

\section{FTS5 vs Semantic Search}
\label{sec:fts5}

This section addresses the core retrieval tradeoff in Cortex: the system is fundamentally anchored in SQLite FTS5 (keyword/BM25 ranking), with additional fallback layers in some retrieval paths. That choice is deliberate and has real consequences.

\textbf{FTS5 advantages:}
\begin{itemize}
  \item Zero external dependencies -- no embedding model, no vector DB, no API call.
  \item Deterministic and auditable: given the same query and index, you get the same results.
  \item Low-latency lookup on typical cortex corpus sizes (hundreds to low thousands of documents).
  \item Works offline, no token cost for retrieval, no rate limits.
\end{itemize}

\textbf{FTS5 limitations:}
\begin{itemize}
  \item Vocabulary mismatch: a query about ``authentication'' won't surface a learning stored as ``login flow.'' Cortex partially mitigates this with a synonym expansion layer, but synonym coverage is hand-curated, not learned.
  \item No semantic relatedness: FTS5 cannot bridge ``the CI pipeline flakes on ARM'' to a learning about ``cross-compilation issues on Apple Silicon,'' even if they describe the same problem.
  \item Ranking quality: BM25 is term-frequency based. It will surface a document that mentions the query word ten times over a more relevant document that mentions it once.
\end{itemize}

\textbf{When the tradeoff bites:} retrieval quality degrades when the user's vocabulary and the stored learning's vocabulary diverge. This is more common in large, long-lived projects with many contributors. It is less common in single-developer projects where vocabulary is consistent.

\textbf{The honest comparison:} semantic search genuinely retrieves better on paraphrase and synonym-heavy queries. Cortex's FTS5 approach retrieves better when the user knows the exact terminology, when offline operation or auditability are required, and when zero retrieval-side token cost matters.

A hybrid approach -- FTS5 for exact matches, semantic or statistical fallback for low-result queries -- closes part of the gap. Cortex implements hybrid retrieval in some paths today, but not uniformly across all entrypoints (see Section~\ref{sec:hybrid}).

\subsection{Hybrid Retrieval: Current State}
\label{sec:hybrid}

Cortex does not currently have one unified hybrid retrieval pipeline. The manual search path (\texttt{search\_knowledge}) and the automatic hook-injection path use different fallback stacks.

Today, the manual search path works as follows:

\begin{enumerate}
  \item \textbf{FTS5 primary}: run the standard BM25 query with synonym expansion.
  \item \textbf{Cosine fallback}: when FTS5 returns few results, compute TF-IDF cosine similarity over a bounded local candidate set.
  \item \textbf{Optional embedding fallback}: when API embeddings are configured, an additional embedding-based fallback can fill remaining result slots.
  \item \textbf{Deduplication}: FTS5 results come first. Cosine results are excluded by rowid to avoid duplicates. FTS5 results are always preferred because they represent exact token matches.
  \item \textbf{Performance guard}: if the corpus exceeds 10,000 documents, the TF-IDF cosine fallback is skipped entirely.
\end{enumerate}

The automatic hook-injection path is more conservative: it stays centered on FTS5 ranking plus a smaller heuristic fallback, with tighter token budgeting and heavier emphasis on determinism.

\textbf{What this means in practice:}

\begin{itemize}
  \item Cortex is not ``FTS-only,'' but it is also not a single, fully unified hybrid system yet.
  \item The strongest current hybrid behavior is in manual search, not automatic prompt injection.
  \item This is a meaningful product distinction because benchmark claims about one path should not automatically be generalized to the other.
\end{itemize}

This design reflects a deliberate choice: a personal project store is not the same retrieval problem as a large enterprise corpus. FTS5 handles the dominant case efficiently, while the fallback layers extend recall where exact vocabulary matching is weak. The remaining design gap is consistency across entrypoints.

\section{When To Choose What}
\subsection{Choose Cortex when}
\begin{itemize}
  \item memory ownership, auditability, and portability are hard constraints,
  \item you need explicit token-budget controls and trust-scored retrieval,
  \item multiple agents share a codebase and need concurrent memory access,
  \item governance matters: role-based access, approval workflows, audit trails,
  \item you want direct memory management through a terminal shell.
\end{itemize}

\subsection{Choose alternatives when}
\begin{itemize}
  \item managed API velocity and external connectors matter more than portability (Supermemory),
  \item deep GitHub-native convenience is primary (Copilot Memory),
  \item local custom hook pipelines with broader runtime composition are preferred (claude-mem),
  \item semantic retrieval quality across large, vocabulary-diverse corpora is critical (Supermemory).
\end{itemize}

\section{Proposed Benchmark Protocol}
\label{sec:benchmark}

\textbf{Note: this is a proposed evaluation framework. No results have been collected against it yet. It is published here to define what a fair comparison would require, not to imply that Cortex has been measured against these criteria.}

A minimum viable benchmark uses:
\begin{itemize}
  \item paired task-level design across systems,
  \item fixed model/runtime controls,
  \item at least 40 tasks across multiple repositories,
  \item bootstrap confidence intervals for primary endpoints.
\end{itemize}

\subsection{Model matrix (practical default)}
Run each system with the same task set across:
\begin{itemize}
  \item Claude family: Opus 4.6, Sonnet 4.6, Haiku 4.6.
  \item OpenAI family: GPT-5.2, GPT-5.3, Codex.
\end{itemize}

Report both \textbf{memory-path latency} (retrieval, ranking, augmentation, persistence) and \textbf{end-to-end latency} (including generation) for each system/model combination.

\subsection{Primary endpoints}
\begin{itemize}
  \item task success rate,
  \item mean injected tokens,
  \item latency p95,
  \item citation validity rate.
\end{itemize}

\section{Limitations}
\label{sec:limitations}

Honest systems documentation acknowledges what doesn't work well. Here is what Cortex gets wrong or trades away:

\textbf{FTS5 retrieval ceiling.} As described in Section~\ref{sec:fts5}, keyword search misses semantically related content. In practice, synonym expansion helps on the most common cases, but vocabulary mismatch remains a real failure mode in large projects.

\textbf{Git contention under concurrent writes.} The Stop hook's commit-and-push design assumes low write concurrency. Under heavy multi-agent parallelism (many agents each finishing responses close together), you will get git merge conflicts or push failures. Cortex does provide in-process write serialization and file locks for MCP writes, but cross-process and cross-machine synchronization still resolves at the git layer.

\textbf{Hook-path indexing overhead.} The FTS5 index is cached, but hook and MCP entrypoints still pay index load and validation costs. For large cortex repos (thousands of markdown files), that adds perceptible latency even when the cache is warm.

\textbf{Ranking quality.} BM25 is not calibrated to coding-context relevance. The quality feedback loop (helpful/reprompt/regression signals) partially compensates over time, but requires enough usage to build signal. New projects start with little or no feedback history, which may surface noisy results early on.

\textbf{Read-path side effects.} In the current implementation, trust filtering can queue stale or conflicting findings during retrieval. This helps surface cleanup work, but it also means some read operations mutate repo state.

\textbf{Governance complexity.} The four-file governance system is powerful but non-trivial to configure correctly. Default settings are reasonable, but teams that need custom TTL, role granularity, or approval workflows will spend time on configuration. There is no guided setup beyond the CLI commands.

\textbf{No web-based review UI by default.} The \texttt{cortex review-ui} command starts a local web UI, but it is not on by default. Memory triage is primarily a terminal workflow. Teams expecting a browser-based management interface will need to adapt.

\textbf{Portability is uni-directional for now.} Export/import works for project data, but migrating the governance configuration (policies, roles, canonical locks) to a different memory system is manual work. The format is not standardized across tools.

\section{Conclusion}
The decision is less about ``who retrieves best on a single prompt'' and more about long-horizon operational fit. Cortex pushes toward production readiness: trust filtering reduces noise in injected context, the governance engine gives teams real policy control, and the interactive shell makes daily memory management possible without starting an LLM session.

Multi-agent support is increasingly non-optional. As coding workflows move from single-agent to team-based coordination, memory systems that assume serial sessions become a bottleneck. Cortex's git-backed architecture provides a workable shared-memory baseline, with the acknowledged caveat that heavy concurrent writes still create contention that requires care.

Managed alternatives remain strong choices where integration speed and platform convenience dominate, or where semantic retrieval quality outweighs portability concerns. The right answer depends on which constraints are binding -- and that answer should be made with eyes open on the tradeoffs documented here.

\FloatBarrier
\bigskip
\small
\sloppy
\begin{thebibliography}{99}
\bibitem{rag2020} Lewis, P. et al. \textit{Retrieval-Augmented Generation for Knowledge-Intensive NLP Tasks}. \url{https://arxiv.org/abs/2005.11401}
\bibitem{selfrag2023} Asai, A. et al. \textit{Self-RAG}. \url{https://arxiv.org/abs/2310.11511}
\bibitem{memorybank2023} Zhong, W. et al. \textit{MemoryBank}. \url{https://arxiv.org/abs/2305.10250}
\bibitem{memgpt2023} Packer, C. et al. \textit{MemGPT}. \url{https://arxiv.org/abs/2310.08560}
\bibitem{sm-intro} Supermemory Docs: Introduction. \url{https://supermemory.ai/docs/introduction}
\bibitem{sm-create} Supermemory Docs: Memory API Creation. \url{https://supermemory.ai/docs/memory-api/creation/adding-memories}
\bibitem{sm-filter} Supermemory Docs: Filtering. \url{https://supermemory.ai/docs/memory-api/features/filtering}
\bibitem{sm-graph} Supermemory Docs: Graph Memory. \url{https://supermemory.ai/docs/concepts/graph-memory}
\bibitem{sm-bench} Supermemory Docs: MemoryBench. \url{https://supermemory.ai/docs/memorybench/quickstart}
\bibitem{cm-overview} claude-mem Docs: Overview. \url{https://docs.claude-mem.ai/architecture/overview}
\bibitem{cm-hooks} claude-mem Docs: Hooks Architecture. \url{https://docs.claude-mem.ai/hooks-architecture}
\bibitem{cm-db} claude-mem Docs: Database Architecture. \url{https://docs.claude-mem.ai/architecture/database}
\bibitem{cm-search} claude-mem Docs: Search Architecture. \url{https://docs.claude-mem.ai/architecture/search-architecture}
\bibitem{gh-memory-docs} GitHub Docs: Copilot Memory. \url{https://docs.github.com/en/copilot/how-tos/use-copilot-agents/copilot-memory}
\bibitem{gh-changelog-jan2026} GitHub Changelog (Jan 15, 2026). \href{https://github.blog/changelog/2026-01-15-agentic-memory-for-github-copilot-is-in-public-preview/}{github.blog changelog link}
\bibitem{gh-changelog-mar2026} GitHub Changelog (Mar 4, 2026). \href{https://github.blog/changelog/2026-03-04-copilot-memory-now-on-by-default-for-pro-and-pro-users-in-public-preview/}{github.blog changelog link}
\bibitem{anthropic-opus46} Anthropic News: Claude Opus 4.6 (Feb 5, 2026). \url{https://www.anthropic.com/news/claude-opus-4-6}
\bibitem{cortex-readme} Cortex README. \url{https://github.com/alaarab/cortex/blob/main/README.md}
\bibitem{cortex-cli} Cortex source: \texttt{mcp/src/cli.ts}. \url{https://github.com/alaarab/cortex/blob/main/mcp/src/cli.ts}
\bibitem{cortex-shared} Cortex source: \texttt{mcp/src/shared.ts}. \url{https://github.com/alaarab/cortex/blob/main/mcp/src/shared.ts}
\bibitem{cortex-hooks} Cortex source: \texttt{mcp/src/hooks.ts}. \url{https://github.com/alaarab/cortex/blob/main/mcp/src/hooks.ts}
\bibitem{cortex-index} Cortex source: \texttt{mcp/src/index.ts}. \url{https://github.com/alaarab/cortex/blob/main/mcp/src/index.ts}
\bibitem{cortex-shell} Cortex source: \texttt{mcp/src/shell.ts}. \url{https://github.com/alaarab/cortex/blob/main/mcp/src/shell.ts}
\bibitem{cortex-data} Cortex source: \texttt{mcp/src/data-access.ts}. \url{https://github.com/alaarab/cortex/blob/main/mcp/src/data-access.ts}
\bibitem{cortex-govern} Cortex source: \texttt{mcp/src/cli-govern.ts}. \url{https://github.com/alaarab/cortex/blob/main/mcp/src/cli-govern.ts}
\bibitem{cortex-extract} Cortex source: \texttt{mcp/src/cli-extract.ts}. \url{https://github.com/alaarab/cortex/blob/main/mcp/src/cli-extract.ts}
\end{thebibliography}
\fussy

\end{document}
